Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-02-07 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flows in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and waterflooding/EOR (enhanced oil recovery) are typical applications in which understandiing the dynamics of immiscible fluid displacement may help mitigating water resources contamination and improving hydrocarbons production. Due to the large viscosity ratio, flow instabilities at fluids' interface may arise leading to formation and growth of fingers, with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers depends mostly on mobility ratio and capillary number, however heterogeneity of the porous domain may affect the onset formation of fingers. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. The model used here is based on a novel control volume finite element method (CVFEM) formulation with a family of FE-pairs, \PN[n]{m} and \PNDG[n]{m} specially tailored for Darcean flows.

  In this work, flow instabilities that arise from fluids' viscosity ratio are numerically investigated.

From initial fluid flow studies of the Hele-Shaw experimental apparatus, a mathematically analogous process to the two-dimensional flow in porous media problem, fingers development as a function of mobility ratio (MR), different domain set-up under uniform or different permeabilities regions and mesh types is examined. A series of cases are listed, showing the validation of the method against experimental setup followed by a more complex 2D domain setup under different viscosity ratios and different mesh resolutions. Finally, a 3D domain setup is showcasing the difference between the preferential flow path and the fingering regimes that will occur in the domain.

%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
For the cases mentioned above, a novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also overcome the computational limitations focusing the effort where is needed, minimizing both the CPU time and the numerical diffusion.

%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Studying the viscous fingering patterns as a function of viscosity ratio along with the initial and boundary conditions, as well as implement adaptivity on top of an unstructured mesh will allow the user to directly apply and setup a geological model accurately and without the need of upscaling. Properties such as porosity and adaptivity can be easily introduced in the model while the idea of the preferential flow path can be distinguished from the finger formation mechanics, in particular the tip-splitting finger behaviour.

>>>>

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flows in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and waterflooding/EOR (enhanced oil recovery) are typical applications in which understandiing the dynamics of immiscible fluid displacement may help mitigating water resources contamination and improving hydrocarbons production. Due to the large viscosity ratio, flow instabilities at fluids' interface may arise leading to formation and growth of fingers -- uneven front with elongation at the outside edge of fluids interface, with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers depends mostly on mobility ratio and capillary number, however heterogeneity of the porous domain may affect the onset formation of fingers. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with a family of FE-pairs, \PN[n]{m} and \PNDG[n]{m} specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio, heterogeneity and mesh resolution in the capture of fluids instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective capture of growth (in particular, tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

%In this work, flow instabilities that arise from fluids' viscosity ratio are numerically investigated. 
%From initial fluid flow studies of the Hele-Shaw experimental apparatus, a mathematically analogous process to the two-dimensional flow in porous media problem, fingers development as a function of mobility ratio (MR), different domain set-up under uniform or different permeabilities regions and mesh types is examined. A series of cases are listed, showing the validation of the method against experimental setup followed by a more complex 2D domain setup under different viscosity ratios and different mesh resolutions. Finally, a 3D domain setup is showcasing the difference between the preferential flow path and the fingering regimes that will occur in the domain.

%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
%For the cases mentioned above, a novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also overcome the computational limitations focusing the effort where is needed, minimizing both the CPU time and the numerical diffusion.

%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
%Studying the viscous fingering patterns as a function of viscosity ratio along with the initial and boundary conditions, as well as implement adaptivity on top of an unstructured mesh will allow the user to directly apply and setup a geological model accurately and without the need of upscaling. Properties such as porosity and adaptivity can be easily introduced in the model while the idea of the preferential flow path can be distinguished from the finger formation mechanics, in particular the tip-splitting finger behaviour.
